---
title: "OmniPath Bioconductor workshop"
author:
  - name: Denes Turei
    affiliation: unihd
    email: turei.denes@gmail.com
    correspondance: true
  - name: Alberto Valdeolivas
    affiliation: unihd
  - name: Attila Gabor
    affiliation: unihd
  - name: Julio Saez-Rodriguez
    affiliation: unihd
institute:
  - unihd: Institute for Computational Biomedicine, Heidelberg University
package: OmnipathR
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
abstract: |
  OmniPath is a database of molecular signaling knowledge, combining data
  from more than 100 resources. It contains protein-protein and gene
  regulatory interactions, enzyme-PTM relationships, protein complexes,
  annotations about protein function, structure, localization and
  intercellular communication. OmniPath focuses on networks with directed
  interactions and effect signs (activation or inhibition) which are suitable
  inputs for many modeling techniques. OmniPath also features a large
  collection of proteins’ intercellular communication roles and interactions.
  OmniPath is distributed by a web service at https://omnipathdb.org/. The
  Bioconductor package OmnipathR is an R client with full support for all
  features of the OmniPath web server. Apart from OmniPath, it provides
  direct access to more than 15 further signaling databases (such as BioPlex,
  InBioMap, EVEX, Harmonizome, etc) and contains a number of convenience
  methods, such as igraph integration, and a close integration with the
  NicheNet pipeline for ligand activity prediction from transcriptomics data.
  In this demo we show the diverse data in OmniPath and the versatile and
  convenient ways to access this data by OmnipathR.
vignette: |
  %\VignetteIndexEntry{OmniPath Bioconductor workshop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
fig_width: 9
fig_height: 7
---

# Introduction

Database knowledge is essential for omics data analysis and modeling. Despite
being an important factor, contributing to the outcome of studies, often
subject to little attention. With [OmniPath](https://omnipathdb.org/) our aim
is to raise awarness of the diversity of available resources and facilitate
access to these resources in a uniform and transparent way. OmniPath has been
developed in a close contact to mechanistic modeling applications and
functional omics analysis, hence it is especially suitable for these fields.
OmniPath has been used for the analysis of various omics data. In the
[Saez-Rodriguez group](https://saezlab.org) we often use it in a pipeline
with our footprint based methods [DoRothEA](
https://saezlab.github.io/dorothea/) and [PROGENy](
https://saezlab.github.io/progeny/) and our causal reasoning method
[CARNIVAL](https://saezlab.github.io/CARNIVAL/) to infer signaling mechanisms
from transcriptomics data.

One recent novelty of OmniPath is a collection of intercellular communication interactions. Apart from simply merging data from existing resources,
OmniPath defines a number of intercellular communication roles, such as
ligand, receptor, adhesion, enzyme, matrix, etc, and generalizes the terms
_ligand_ and _receptor_ by introducing the terms _transmitter_, _receiver_
and _mediator_. This unique knowledge base is especially adequate for the
emerging field of cell-cell communication analysis, typically from single
cell transcriptomics, but also from other kinds of data.

# Overview

## Pre-requisites

No special pre-requisites apart from basic knowledge of R. OmniPath, the
database resource in the focus of this workshop has been published in [1,2],
however you don't need to know anything about OmniPath to benefit from the
workshop. In the workshop we will demonstrate the [R/Bioconductor package](
https://bioconductor.org/packages/release/bioc/html/OmnipathR.html)
[OmnipathR](https://github.com/saezlab/OmnipathR). If you would like to try
the examples yourself we recommend to install the latest version of the
package before the workshop:

```{r, eval=FALSE}
library(devtools)
install_github('saezlab/OmnipathR')
```

## Participation

In the workshop we will present the design and some important features of the
OmniPath database, so can be confident you get the most out of it. Then we
will demonstrate further useful features of the OmnipathR package, such as
accessing other resources, building graphs. Participants are encouraged to
experiment with the examples and shape the contents of the workshop by asking
questions. We are happy to recieve questions and topic suggestions **by
email** also **before the workshop**. These could help us to adjust the
contents to the interests of the participants.

## _R_ / _Bioconductor_ packages used

* OmnipathR
* igraph
* dplyr

## Time outline

Total: 45 minutes

| Activity                     | Time |
|------------------------------|------|
| OmniPath database overview   | 5m   |
| Network datasets             | 10m  |
| Other OmniPath databases     | 5m   |
| Intercellular communication  | 10m  |
| Igraph integration           | 5m   |
| Further resources            | 10m  |

## Workshop goals and objectives

In this workshop you will get familiar with the design and features of the
OmniPath databases. For example, to know some important details about the
datasets and parameters which help you to query the database the most
suitable way according to your purposes. You will also learn about
functionalities of the _OmnipathR_ package which might make your work
easier.

## Learning goals

* Learn about the OmniPath database, its contents and how it can be useful
* Get a picture about the OmnipathR package capabilities
* Learn about the datasets and parameters of various OmniPath query types

## Learning objectives

* Try examples of each OmniPath query type with various parameters
* Build igraph networks, search for paths
* Access some further interesting resources

# Workshop

Here we will publish the workshop contents with code examples.

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References {.unnumbered}

[1] D Turei, A Valdeolivas, L Gul, N Palacio-Escat, O Ivanova, A Gabor,
D Modos, T Korcsmaros and J Saez-Rodriguez (2020) Integrated intra- and
intercellular signaling knowledge for multicellular omics analysis. _bioRxiv_
2020.08.03.221242

[2] D Turei, T Korcsmaros and J Saez-Rodriguez (2016) OmniPath: guidelines and
gateway for literature-curated signaling pathway resources. _Nature Methods_
13(12)