---
title: "Building prior knowledge network (PKN) for COSMOS"
author:
  - name: Denes Turei
    email: turei.denes@gmail.com
    correspondance: true
    affiliation: Institute for Computational Biomedicine, Heidelberg University
institute:
  - unihd: Institute for Computational Biomedicine, Heidelberg University
package: OmnipathR
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
abstract: |
  The prior knowledge network (PKN) used by COSMOS is a network of
  heterogenous causal interactions: it contains protein-protein,
  reactant-enzyme and enzyme-product interactions. It is a combination of
  multiple resources. Here we present the functions that load each component,
  the options for customization, and the functions to build the complete PKN.
  \itemize{
  }
vignette: |
  %\VignetteIndexEntry{COSMOS PKN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
fig_width: 9
fig_height: 7
---

# Introduction

The COSMOS PKN is a combination of the following datasets:

* Genome-scale metabolic model (GEM) from Chalmers Sysbio (Wang et al., 2021.)
* Network of chemical-protein interactions from STITCH (http://stitch.embl.de/)
* Protein-protein interactions from Omnipath (Türei et al., 2021)

Building the PKN is possible in by calling the `cosmos_pkn()` function that we
present in the last section. First let's take a closer look at each resource.

```{r message=FALSE, warning=FALSE}
library(OmnipathR)
```

# Chalmers Sysbio GEM

```{r chalmers}
ch <- chalmers_gem_network()

ch
```

# STITCH enzyme-metabolite interactions

```{ri stitch}
st <- stitch_gem()

st
```

# OmniPath signaling network

```{r omnipath}
op <- omnipath_for_cosmos()

op
```

# Complete build

```{r cosmos}
pkn <- cosmos_pkn()

pkn
```

The `record_id` column identifies the original records within each
resource. If one `record_id` yields multiple records in the final data
frame, it is the result of one-to-many ID translation or other
processing steps. Before use, it is recommended to select one pair of ID
type columns (by combining the preferred ones) and perform `distinct`
by the identifier columns and sign.
