---
title: "The database manager in OmnipathR"
author:
  - name: Denes Turei
    email: turei.denes@gmail.com
    correspondance: true
    affiliation: Institute for Computational Biomedicine, Heidelberg University
institute:
  - unihd: Institute for Computational Biomedicine, Heidelberg University
package: OmnipathR
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
abstract: |
  The database manager is an API within OmnipathR which is able to load
  various datasets, keep track of their usage and remove them after an
  expiry period. Currently it supports a few Gene Ontology and UniProt
  datasets, but easily can be extended to cover all datasets in the
  package.
vignette: |
  %\VignetteIndexEntry{Database manager}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
fig_width: 9
fig_height: 7
---

# Available datasets

To see a full list of datasets call the `omnipath_show_db` function:

```{r pipe, include=FALSE}
`%>%` <- magrittr::`%>%`
```

```{r show}
library(OmnipathR)
omnipath_show_db()
```

It returns a tibble where each dataset has a human readable name and a key
which can be used to refer to it. We can also check here if the dataset is
currently loaded, the time it's been last used, the loader function and its
arguments.

# Access a dataset

Datasets can be accessed by the `get_db` function. Ideally you should call
this function every time you use the dataset. The first time it will be
loaded, the subsequent times the already loaded dataset will be returned.
This way each access is registered and extends the expiry time. Let's load
the human UniProt-GeneSymbol table. Above we see its key is `up_gs`.

```{r get}
up_gs <- get_db('up_gs')
up_gs
```

This dataset is a two columns data frame of SwissProt IDs and Gene Symbols.
Looking again at the datasets, we find that this dataset is loaded now and
the `last_used` timestamp is set to the time we called `get_db`:

```{r show-2}
omnipath_show_db()
```

The above table contains also a reference to the dataset, and the arguments
passed to the loader function:

```{r show-3}
d <- omnipath_show_db()
d %>% dplyr::pull(db) %>% magrittr::extract2(16)
d %>% dplyr::pull(latest_param) %>% magrittr::extract2(16)
```

If we call `get_db` again, the timestamp is updated, resetting the expiry
counter:

```{r sleep, include=FALSE}
# some sleep time to make sure we see a time difference
Sys.sleep(10)
```

```{r expiry}
up_gs <- get_db('up_gs')
omnipath_show_db()
```

# Where are the loaded datasets?

The loaded datasets live in an environment which belong to the OmnipathR
package. Normally users don't need to access this environment. As we see
below, `omnipath_show_db` presents us all information availble by directly
looking at the environment:

```{r env}
OmnipathR:::omnipath.env$db$up_gs
```

# How to extend the expiry period?

The default expiry of datasets is given by the option `omnipath.db_lifetime`.
By calling `omnipath_save_config` this option is saved to the default config
file and will be valid in all subsequent sessions. Otherwise it's valid only
in the current session.

```{r lifetime, eval=FALSE}
options(omnipath.db_lifetime = 600)
omnipath_save_config()
```

# Where are the datasets defined?

The built-in dataset definitions are in a JSON file shipped with the package.
Easiest way to see it is by [the git web interface][1].

# How to add custom datasets?

Currently no API available for this, but it would be super easy to implement.
It would be matter of providing a JSON similar to the above, or calling a
function. Please open an issue if you are interested in this feature.

# Session information

```{r session}
sessionInfo()
```

[1]: https://github.com/saezlab/OmnipathR/blob/master/inst/db/db_def.json
