---
title: "The omnipath-metabo Web Service Client"
author:
    - name: OmniPath Team
output:
    BiocStyle::html_document:
        number_sections: yes
        toc: yes
        toc_depth: 4
        pandoc_args:
            - '--lua-filter=scholarly-metadata.lua'
            - '--lua-filter=author-info-blocks.lua'
vignette: >
    %\VignetteIndexEntry{The omnipath-metabo Web Service Client}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

# Overview

[omnipath-metabo](https://github.com/saezlab/omnipath-metabo) is a Python
package that builds and serves metabolite, compound and reaction-related
prior-knowledge, most notably the **COSMOS PKN** (prior-knowledge network
for multi-omics causal reasoning) and a growing set of other **specialized
networks** (e.g. MetalinksDB, LIANA). It exposes this data over a REST API
at `https://metabo.omnipathdb.org`.

The `metabo_*` functions in OmnipathR are a thin, cached client for that
web service. They are **unrelated** to `cosmos_pkn()` (see
`vignette("cosmos")`), which builds a PKN locally from STITCH, the Chalmers
GEM and OmniPath -- the two do not share code or output, and using one does
not preclude using the other.

```{r eval=FALSE}
library(OmnipathR)
```

# Discovering what's available

The set of PKN categories, organisms and contributing resources reflects
whatever the service currently has built and cached -- nothing here is
hard-coded in the package, so these calls always return the current state
of the service:

```{r eval=FALSE}
metabo_cosmos_categories()
#> [1] "transporters"      "receptors"         "allosteric"
#> [4] "enzyme_metabolite" "ppi"               "grn"

metabo_cosmos_organisms()
#> [1]  9606 10090 10116

metabo_cosmos_resources()
#> # A tibble: 12 x 2
#>   category     resource
#>   <chr>        <chr>
#> 1 allosteric   BRENDA
#> ...
```

# Retrieving a COSMOS PKN subset

`metabo_cosmos_pkn()` retrieves the PKN for one organism, optionally
restricted to one or more categories and/or contributing resources:

```{r eval=FALSE}
# One category
transporters <- metabo_cosmos_pkn(organism = 9606, categories = "transporters")

# Several categories
metabolic <- metabo_cosmos_pkn(
    organism = 9606,
    categories = c("enzyme_metabolite", "transporters")
)

# The full PKN (default)
full_pkn <- metabo_cosmos_pkn(organism = "human")
```

Every interaction carries `source`, `target`, an effect sign (`mor`), the
contributing `resource`, and an `attrs` list column preserving
resource-specific metadata (such as `reaction_id` for metabolic-reaction
edges) for future use -- see `?metabo_cosmos_pkn` for the full column
reference.

Results are cached locally through OmnipathR's existing download cache, so
repeating an identical query does not re-hit the network:

```{r eval=FALSE}
# First call downloads and caches
metabo_cosmos_pkn(organism = 9606, categories = "transporters")

# Second, identical call is served from the local cache
metabo_cosmos_pkn(organism = 9606, categories = "transporters")
```

# Combined, purpose-named views

Two convenience wrappers cover the most common combinations without
needing to assemble a category list by hand:

```{r eval=FALSE}
# The complete network connecting metabolism to signalling and gene
# regulation -- equivalent to categories = "all"
signaling_pkn <- metabo_metabolic_signaling_pkn(organism = 9606)

# Metabolite-protein interactions only (receptor + allosteric-regulation
# categories) -- comparable in scope to a resource like MetalinksDB, but
# derived from the COSMOS PKN build itself
metalinks_like <- metabo_metabolite_protein_pkn(organism = 9606)
```

For any other combination, call `metabo_cosmos_pkn()` directly with the
`categories` you need.

# Querying any specialized network

Beyond the COSMOS PKN, the service registers other specialized networks
(e.g. MetalinksDB, LIANA). `metabo_networks()` lists what is currently
registered -- again, discovered at call time, not hard-coded:

```{r eval=FALSE}
metabo_networks()
#> # A tibble: 2 x 6
#>   name        kind             schema_name  combined_relation included_sources built_at
#>   <chr>       <chr>            <chr>        <chr>             <list>           <chr>
#> 1 liana       ligand_receptor  custom_views liana_ligand_re...  <chr [5]>       2026-06-...
#> 2 metalinksdb compound_protein custom_views metalinksdb_rel...  <chr [12]>      2026-06-...
```

Retrieve one page of interactions, or page through the whole network:

```{r eval=FALSE}
metabo_network_status("metalinksdb")
metabo_network_resources("metalinksdb")

# One page
page <- metabo_network_interactions("metalinksdb", limit = 1000)

# The complete network, paged automatically
all_rows <- metabo_network_interactions("metalinksdb", paginate = TRUE)
```

The column set of the returned table is not fixed by OmnipathR -- it
reflects whatever the requested network exposes, so new networks are
served correctly without a package update.

Requesting an unknown network name raises a clear error pointing back at
`metabo_networks()`, instead of a raw HTTP failure:

```{r eval=FALSE}
metabo_network_interactions("not-a-real-network")
#> Error: Unknown network: 'not-a-real-network' (use `metabo_networks()` to
#> list valid names).
```

# Configuration

Like the main OmniPath API client, the omnipath-metabo base URL is
configurable, e.g. to point at a development or staging deployment:

```{r eval=FALSE}
omnipath_get_option("omnipathr.metabo_url")
#> [1] "https://metabo.omnipathdb.org/"

omnipath_set_option(omnipathr.metabo_url = "https://metabo-dev.omnipathdb.org/")
```

It can also be set via the `OMNIPATHR_METABO_URL` environment variable.

# Current limitations

- Only the binary (flat source-target edge list) PKN representation is
  supported. A reaction-grouped ("hypergraph") representation, collapsing
  rows that share the same `attrs$reaction_id` into one multi-substrate
  hyperedge, is not yet implemented.
- Only JSON transfer is supported; the service's Parquet response format
  is not used by this client.
