---
title: "Omnipath R interface"
author: "Attila Gabor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Omnbipath R interface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(OmnipathR)
```


## Omnipath R

This package is inteded to serve as a light interface to 
[Omnipath Database](http://www.omnipathdb.org).
In it's simplest form the only thing it does, is to download all the interactions
as a `data.frame` and optionally converts it to an `igraph` object. 


## Databases

Omnipath is a collection of many databases (comprehensive list)[http://www.omnipathdb.org/info]. 
Please browse [Omnipath Database](http://www.omnipathdb.org) to find the currently available databases. 
As now the following interactions are available:  
- Interactions: contains protein protein interactions, possibly signed and directed.  
- PTMS: post-translational modifications of protein interactions  
- Complexes (to be implemented)
- Annotations (to be implemented)


## Basic functions

Download the interaction databases and select some of the 
```{r}
ptms = import_Omnipath_PTMS(filter_databases=c("PhosphoSite", "Signor"))
print(head(ptms))

interactions = import_Omnipath_Interactions(filter_databases=c("SignaLink3","PhosphoSite", "Signor"))
print(head(interactions))
```

We can also convert them to an igraph object: 
```{r}

ptms_g = ptms_graph(ptms = ptms )
ptms_g
OPI_g = interaction_graph(interactions = interactions )
OPI_g
```



Further, we provide some utility functions to browse 
